Re-examining cross-cultural similarity judgments using lexical co-occurrence

### 2023-02-01 code/4_paper/manuscript/manuscript.tex
manuscript edits

\author{{\large \bf Khuyen N. Le (knl005@ucsd.edu)}$^{1\star}$, {\large \bf Alexandra Carstensen (abc@ucsd.edu)}$^{1\star}$, \\ {\large \bf Shan Gao (shangaocog@gmail.com)$^2$} {\large \bf Michael C. Frank (mcfrank@stanford.edu)$^3$},  \AND $^1$Department of Psychology, UC San Diego, $^2$Department of Psychology, University of Chicago, \\ $^3$Department of Psychology, Stanford University }

@@

One hypothesis is that differences in similarity judgment is linked to
the tendency to see more analytic processing in Western culture and more
holistic processing in East Asian cultures (Nisbett (2003)). On this
hypothesis, since analytic processing emphasizes relationships between
objects and their properties, it leads to taxonomic similarity
judgments; in contrast, holistic processing emphasizes relations between
objects and leads to thematic similarity judgments. Cross-cultural
difference in similarity judgment is thus connected to work showing East
Asian participants exhibiting a higher level of sensitivity to context
than their Western counterparts when reproducing drawings from memory
(Ji, Peng, \& Nisbett, 2000); visually exploring naturalistic scenes
(Chua, Boland, \& Nisbett, 2005); describing scenes (Masuda \& Nisbett,
2001); and in explaining the causes of ambiguous behaviors (Choi,
Nisbett, \& Norenzayan, 1999). This view ascribes cross-cultural
differences in similarity judgment to variation in how each culture (or
group of culture, i.e.~East Asian vs Western) \emph{conceptualize}
similarity -- with East Asians preferring thematic relations more than
Westerners.

@@

\hypertarget{estimating-variation-in-experience}{%
\subsection{Estimating variation in
experience}\label{estimating-variation-in-experience}}

To test alternative hypotheses involving variation of inputs to
similarity judgment, we need a way to operationalize variation in
exposure to thematic vs taxonomic relations. While this metric is
difficult to directly measure, language statistics can provide a rough
proxy. Language statistics is useful in that it is part of the inputs to
everyday experiences (through spoken and written language), may afford
many of the `experiences' that people have with infrequently encountered
items, like cows or helicopters, and provides accessible measures.
Previous work also suggests that using language statistics (such as
lexical co-occurrence or cosine distance of word embeddings) as a proxy
can be a good predictor for similarity reasoning. A model trained on
word-document co-occurrence can predict word association and the effects
of semantic association on a variety of linguistic tasks (Griffiths,
Steyvers, \& Tenenbaum, 2007). Lexical semantic models that are
constructed using lexical co-occurrence (in comparison to annotated
relations) have been shown to perform well on predicting human judgments
about similarity between word pairs that are thematically or
taxonomically related (Rohde, Gonnerman, \& Plaut, 2006). Word
embeddings like word2vec, gloVe, and fastText have been demonstrated to
be good predictors for similarity judgments (Liu, Feng, Wu, Chan, \&
Fulton (2019), Jatnika, Bijaksana, \& Suryani (2019)).

@@

We note that language statistics can incorporate both cultural-specific
environmental statistics (the inputs of similarity judgment) and
taxonomic/thematic preferences (the conceptualization of similarity).
For example, cultural features (farming) can lead to differences in
environmental statistics (seeing cows and grass) and in turn these can
influence language (talking more about cows and grass). But cultural
features (preferring thematic relations) could also more directly cause
individuals to talk differently about the same experiences (mentioning
what cows eat rather than what other animals cows are like). Therefore,
our approach looks at the extent to which language statistics can
predict cross-cultural differences in similarity judgment with the
understanding that language statistics is a proxy for both inputs and
conceptualization of similarity judgment\footnote{However, in Q3, we
  investigate whether differing culture-based conceptualization of
  similarity has unique contribution to cross-cultural variation of
  similarity jdugment.}.

@@

Q4: Is the relationship with co-occurrence specific to
taxonomic/thematic items, or can co-occurrence structure predict other
judgments?~

To preview our results, we replicate US-China differences and find that
Vietnamese judgments are intermediate between these two. Further,
language-specific statistics are a good predictor for cross-cultural
differences in similarity judgments. Language statistics are also a good
predictor for filler items with no taxonomic/thematic structure. Our
findings are in contrast to previous accounts, by which culture induces
differing conceptions of similarity (varying between taxonomic and
thematic). these findings provide an alternative account by which
language may explain these cross-cultural differences without invoking
variation in notions of similarity.

@@

We adapted stimuli from previous studies to create a set of test triads
consisting of a cue, with one thematic and one taxonomic match option.
For example, ``cow,'' ``grass,'' and ``chicken,'' where ``cow'' is the
cue, ``grass'' is the thematic match, and ``chicken'' the taxonomic
match. We included 105 such triads, a superset including triads pulled
from supplemental information and in-text examples across the
literature, and others that we adapted or created . We selected triads
on the basis of cultural familiarity in the US, Vietnam, and China. The
triads were originally in English; they were translated to Vietnamese
and Mandarin by a fluent bilingual speaker in each language. The
translations were then checked for accuracy after backtranslation to
English by another fluent bilingual in each language who was naive to
the original English versions. All materials are included in the
Supplementary Information.

@@

In practice, we calculated the cosine distance between each cue-thematic
match (thematic cosine distance) and cue-taxonomic match (taxonomic
cosine distance), using the spatial.distance.cosine function from the
SciPy package (Virtanen et al., 2020). We then calculated the thematic
cosine distance proportion as thematic cosine distance over the sum of
taxonomic cosine distance and thematic cosine distance. We did this for
all three corpora. We were able to obtain predictions for all triads in
all languages. We then use a mixed-effects regression to evaluate how
well each corpus model predicts participants' similarity judgments,
across triads and cultural contexts.

\hypertarget{results}{%
\section{Results}\label{results}}

@@

\hypertarget{cultural-context-as-a-predictor-for-cross-cultural-variation-in-similarity-judgments}{%
\subsection{3. Cultural context as a predictor for cross-cultural
variation in similarity
judgments}\label{cultural-context-as-a-predictor-for-cross-cultural-variation-in-similarity-judgments}}

Our above models tested whether language statistics is a good predictor
for similarity judgment. As we noted above, language statistics can be a
proxy of both inputs and conceptualization of similarity. However, it is
still an open question whether conceptualization of similarity itself
has a unique contribution to similarity judgment (beyond its influence
on language statistics). To test this hypothesis, we operationalize
cultural-specific conceptualization of similarity as country. We then
compare a model that includes country, corresponding corpus statistics
and their interaction as fixed effects to a model with only the
corresponding corpus statistics as the fixed effect. If
conceptualization of similarity has a unique contribution to similarity
judgment, we should see that terms including country to be a significant
predictor of responding.

In the model containing only the corresponding corpus statistics, the
corpus statistics is a significant predictor (\(\beta\) = -1.79,
\(\chi^2\)(1) = 75.58, p \textless{} .001). When adding country and
interaction between country and corpus, corpus statistics (\(\beta\) =
-1.72, \(\chi^2\)(1) = 23.34, p \textless{} .001), country (F(2,
\ensuremath{\infty{}}) = 13.18, \(\chi^2\)(2) = 7.39, p = 0.025), and
interaction between country and corpus (F(2, \ensuremath{\infty{}}) =
6.66, \(\chi^2\)(2) = 13.33, p = 0.001), are significant
predictors\footnote{With preregistered exclusion criterion: only corpus
  statistics and interaction between country and corpus are significant
  predictors.}. Including the country and country-corpus terms to the
language-only model significantly its ability to explain variation in
responding (\(\chi^2\)(4) = 40.76, p \textless{} .001). This result
points to a unique contribution by culture-specific ways of
conceptualizing similarity to similarity judgment that is not captured
in language statistics.

\hypertarget{language-statistics-as-a-predictor-for-non-structured-triad-items}{%
\subsection{4. Language statistics as a predictor for non-structured
triad
items}\label{language-statistics-as-a-predictor-for-non-structured-triad-items}}

A possible concern with our approach is that variation in language
statistics (which we claim to be a proxy of both inputs and
conceptualization of similarity) can be fully explained by variation in
conceptualization of similarity (i.e.~the difference in similarity
measures between `cow' and `grass' versus `cow' and `chicken' in one
corpus is solely driven by the corresponding culture's preference of
thematic versus taxonomic categorization). Alternatively, language
statistics may be measuring slight but pervasive differences in
similarity driven by participants' different cultural contexts, rather
than merely the type of similarity they generally prefer (thematic vs
taxonomic relations). We address this concern by testing whether
language statistics covary with similarity judgment of items with no
taxonomic/thematic structures (filler items). If language statistics is
only driven by conceptualization of similarity, we should not see an
effect with such items, which have no structured similarity relations by
which to make any predictions.~

@@

Using the same mixed-effects logistic regression structure as Q2, we
predict responses (1=word1 or 0=word2) of each cultural context with the
corresponding corpus statistics as a fixed effect, and participant and
triad as random effects. We found that in all cultural contexts, the
corresponding corpus is a significant predictor of responding, with p
\textless{} 0.05 and \(\beta\) from -9.59 to -3.14. (For a full report,
see Supplementary Information.)

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

In this paper, we consider whether statistics of the environment (as
proxied by language statistics) can account for cross-cultural
differences in a classic similarity judgment paradigm, as an alternative
to the view that members of different cultures vary in their
conceptualization of similarity. While we managed to replicate the
previously documented contrast between English speakers in the US and
Mandarin Chinese speakers from East Asia (mainland China, Taiwan, Hong
Kong, and Singapore), we do not extend this contrast to our sample of
Vietnamese speakers in Vietnam and English speakers in the US. This
finding suggests some limitations on the generality of the cultural
account. We found some support for the environmental statistics account:
each corpus statistics is a good predictor for the corresponding
country's similarity judgments, even when other corpus statistics are
included, and even with triads without a taxonomic/thematic structure.
However, we also found that culture-based conceptualization of
similarity uniquely contributes to predicting similarity judgment when
included in a model with corpus statistics. Overall, our results provide
evidence that cross-cultural differences in similarity judgment are
related to linguistic co-occurrence patterns (which also vary across
cultures).

@@

Our findings raise additional questions for future work: To what extent
are the relativity effects driven by language, and to what extent by
culture? Ji et al. (2004) established that culture-aligned differences
in this paradigm exist, even when the test language is held constant,
concluding that ``it is culture (independent of the testing language)
that led to different grouping styles'' in their study. Our data provide
a cautionary note to this conclusion, suggesting that semantic
representations in bilinguals (see Francis (2005) for a review) may have
the potential to provide an offline account for cross-context
differences in similarity judgments, independent of test language.
However, there are still many open questions for this account. How do
semantic associations guide categorization? Can they explain
taxonomic-thematic differences of the type reported by Ji et al. (2004)
and others? Can we provide a more specific computational account than
the simple proportion-of-similarity model tested here?

@@

\leavevmode\vadjust pre{\hypertarget{ref-Francis2005}{}}%
Francis, W. S. (2005). \emph{Bilingual semantic and conceptual
representation}. Oxford University Press.

>>>>

\author{{\large \bf Khuyen N. Le (knl005@ucsd.edu)}$^{1}$, {\large \bf Alexandra Carstensen (abc@ucsd.edu)}$^{1}$, \\ {\large \bf Shan Gao (shangaocog@gmail.com)$^2$}, {\large \bf Michael C. Frank (mcfrank@stanford.edu)$^3$},  \AND $^1$Department of Psychology, University of California, San Diego, $^2$Department of Psychology, University of Chicago, \\ $^3$Department of Psychology, Stanford University }

@@

One hypothesis is that differences in similarity judgment is linked to
the tendency to see more analytic processing in Western culture and more
holistic processing in East Asian cultures (Nisbett, 2003). On this
hypothesis, since analytic processing emphasizes relationships between
objects and their properties, it leads to taxonomic similarity
judgments; in contrast, holistic processing emphasizes relations between
objects and leads to thematic similarity judgments. Cross-cultural
difference in similarity judgment is thus connected to work showing East
Asian participants exhibiting a higher level of sensitivity to context
than their Western counterparts when reproducing drawings from memory
(Ji, Peng, \& Nisbett, 2000); visually exploring naturalistic scenes
(Chua, Boland, \& Nisbett, 2005); describing scenes (Masuda \& Nisbett,
2001); and in explaining the causes of ambiguous behaviors (Choi,
Nisbett, \& Norenzayan, 1999). This view ascribes cross-cultural
differences in similarity judgment to variation in how each culture (or
group of culture, i.e.~East Asian vs Western) \emph{conceptualize}
similarity -- with East Asians preferring thematic relations more than
Westerners.

@@

\hypertarget{estimating-variation-in-experience-via-language-statistics}{%
\subsection{Estimating variation in experience via language
statistics}\label{estimating-variation-in-experience-via-language-statistics}}

To test alternative hypotheses involving variation of inputs to
similarity judgment, we need a way to operationalize variation in
exposure to thematic vs taxonomic relations. While this metric is
difficult to directly measure, language statistics can provide a rough
proxy. Language statistics is useful in that it is part of the inputs to
everyday experiences (through spoken and written language), may afford
many of the `experiences' that people have with infrequently encountered
items, like cows or helicopters, and provides accessible measures.
Previous work also suggests that using language statistics (such as
lexical co-occurrence or cosine distance of word embeddings) as a proxy
can be a good predictor for similarity reasoning. A model trained on
word-document co-occurrence can predict word association and the effects
of semantic association on a variety of linguistic tasks (Griffiths,
Steyvers, \& Tenenbaum, 2007). Lexical semantic models that are
constructed using lexical co-occurrence (in comparison to annotated
relations) have been shown to perform well on predicting human judgments
about similarity between word pairs that are thematically or
taxonomically related (Rohde, Gonnerman, \& Plaut, 2006). Word
embeddings like word2vec, gloVe, and fastText have been demonstrated to
be good predictors for similarity judgments (Jatnika, Bijaksana, \&
Suryani, 2019; Liu, Feng, Wu, Chan, \& Fulton, 2019).

@@

We note that language statistics can incorporate both cultural-specific
environmental statistics (the inputs of similarity judgment) and
thematic/taxonomic preferences (the conceptualization of similarity).
For example, cultural features (farming) can lead to differences in
environmental statistics (seeing cows and grass) and in turn these can
influence language (talking more about cows and grass). But cultural
features (preferring thematic relations) could also more directly cause
individuals to talk differently about the same experiences (mentioning
what cows eat rather than what other animals cows are like). Therefore,
our approach looks at the extent to which language statistics can
predict cross-cultural differences in similarity judgment with the
understanding that language statistics is a proxy for both inputs and
conceptualization of similarity judgment\footnote{However, in Q3, we
  investigate whether differing culture-based conceptualization of
  similarity has unique contribution to cross-cultural variation of
  similarity jdugment.}.

@@

Q4: Is the relationship with co-occurrence specific to
thematic/taxonomic items, or can co-occurrence structure predict other
judgments?~

To preview our results, we replicate US-China differences and find that
Vietnamese judgments are intermediate between these two. Further,
language-specific statistics are a good predictor for cross-cultural
differences in similarity judgments. Language statistics are also a good
predictor for filler items with no thematic/taxonomic structure. Our
findings are in contrast to previous accounts, by which culture induces
differing conceptions of similarity (varying between taxonomic and
thematic). these findings provide an alternative account by which
language may explain these cross-cultural differences without invoking
variation in notions of similarity.

@@

We adapted stimuli from previous studies to create a set of test triads
consisting of a cue, with one thematic and one taxonomic match option.
For example, ``cow,'' ``grass,'' and ``chicken,'' where ``cow'' is the
cue, ``grass'' is the thematic match, and ``chicken'' the taxonomic
match. We included 105 such triads, a superset including triads pulled
from supplemental information and in-text examples across the
literature, and others that we adapted or created . We selected triads
on the basis of cultural familiarity in the US, Vietnam, and China. The
triads were originally in English; they were translated to Vietnamese
and Mandarin by a fluent bilingual speaker in each language. The
translations were then checked for accuracy after backtranslation to
English by another fluent bilingual in each language who was naive to
the original English versions. All materials are available at {[}BLINDED
FOR REVIEW{]}.

@@

In practice, we calculated the cosine distance between each cue-thematic
match (thematic cosine distance) and cue-taxonomic match (taxonomic
cosine distance), using the spatial.distance.cosine function from the
SciPy package (Virtanen et al., 2020). We then calculated the thematic
cosine distance proportion as thematic cosine distance over the sum of
taxonomic cosine distance and thematic cosine distance. We did this for
all three corpora. We were able to obtain predictions for all triads in
all languages. We then use a mixed-effects regression to evaluate how
well each corpus model predicts participants' similarity judgments,
across triads and cultural contexts. All analyses are available at
{[}BLINDED FOR REVIEW{]}.

\hypertarget{results-discussion}{%
\section{Results \& Discussion}\label{results-discussion}}

@@

\hypertarget{language-statistics-as-a-predictor-for-non-structured-filler-items}{%
\subsection{3. Language statistics as a predictor for non-structured
filler
items}\label{language-statistics-as-a-predictor-for-non-structured-filler-items}}

A possible concern with our approach is that our models using language
statistics in Q2 are only able to account for variation in
thematic/taxonomic preference in similarity judgment because language
statistics is picking up on differences in thematic/taxonomic preference
in conceptualization of similarity. For example, perhaps ``cow'' and
``grass'' co-occurs more than ``cow'' and ``chicken'' in Mandarin
because Chinese speakers prefer thematic relations, and thus talk about
``cow'' and ``grass'' together more often. Vice versa, English speakers
preferring taxonomic relations would talk about ``cow'' and ``chicken''
together more often, hence driving higher lexical co-occurrence for
``cow'' and ``chicken'' in the English corpus. In other words, perhaps
the language statistics models are merely measuring variation in
thematic/taxonomic relation preference in conceptualization of
similarity, only in a different way. If this is the case, language
statistics should not be able to predict responding for filler items
(such as whether ``tomorrow'' or ``yesterday'' is more related to
``today'') because these items do not have a thematic/taxonomic
structure. Additionally, even when one option might be more related to
the cue, such relationship is not systematic throughout the set of
fillers. On the other hand, if language statistics capture similarity
judgment in a more general sense (beyond thematic/taxonomic preference
in conceptualization of similarity), it should also be a significant
predictor for the non-structured filler items.

@@

Using the same mixed-effects logistic regression structure as the single
corpus models in Q2, we predict responses (1=word1 or 0=word2) of each
cultural context with the corresponding corpus statistics as a fixed
effect, and participant and triad as random effects. We found that in
all cultural contexts, the corresponding corpus is a significant
predictor of responding, with p \textless{} 0.05 and \(\beta\) from
-9.59 to -3.14. (For a full report, see Supplementary Information.)
These results show that the language statistics model is accounting for
general similarity in addition to structured thematic/taxonomic
similarity.

To investigate whether language statistics predicts triads and filler
items differently, we compare the variance in responding that can be
accounted for by language statistics of the corresponding corpus in the
two different types of items (structured triads versus non-structured
filler items). Namely, we ran two mixed-effects logistic regression
models that predict responding across cultural contexts (1=thematic or
word1, 0=taxonomic or word2), with the corresponding corpus statistics
(for e.g., cosine proportion from the English corpus for US responding)
as a fixed effect and participant and item as random effects. Consistent
with our results above, corresponding corpus is a significant predictor
for responding to both triads and filler items (triads: \(\beta\) =
-1.79, \(\chi^2\)(1) = 75.58, p \textless{} .001, fillers: \(\beta\) =
-2, \(\chi^2\)(1) = 70.89, p \textless{} .001). Importantly, we found
comparable conditional \(R^2\) when corresponding corpus is used to
predict only triad items (\(R^2\) = 0.3) and only filler items (\(R^2\)
= 0.26), suggesting that culture-corresponding corpus statistics is able
to explain the same amount of variance in the responding data for either
triads or filler items. This result provides evidence for our view that
language statistics is measuring general tendencies in similarity
judgment, beyond preference for thematic/taxonomic relations.

\hypertarget{general-discussion}{%
\section{General Discussion}\label{general-discussion}}

In this paper, we consider whether statistics of the environment (as
proxied by language statistics) can account for cross-cultural
differences in a classic similarity judgment paradigm, as an alternative
to the view that members of different cultures vary in their
conceptualization of similarity. We replicated the previously documented
contrast between English speakers in the US and Mandarin Chinese
speakers from East Asia (mainland China, Taiwan, Hong Kong, and
Singapore), with mainland China participants preferring thematic
relations to a greater extent compared to US participants. Our sample of
Vietnamese participants showed intermediate response but not
significantly different from either Chinese or US participants. This
finding suggests some limitations on the generality of the cultural
account, which proposes that thematic/taxonomic similarity preference
aligns with East Asian/Western tendency for holistic vs analytic
processing. We found some support for the environmental statistics
account: each corpus statistics is a good predictor for the
corresponding country's similarity judgments, even when other corpus
statistics are included, and even with triads without a
thematic/taxonomic structure. Overall, our results provide evidence that
cross-cultural differences in similarity judgment are related to
linguistic co-occurrence patterns (which also vary across cultures).

@@

Ji et al. (2004) established that culture-aligned differences in this
paradigm exist, even when the test language is held constant, concluding
that ``it is culture (independent of the testing language) that led to
different grouping styles'' in their study. Our data provide a
cautionary note to this conclusion, suggesting that while cultural
differences in similarity judgment exists, we might get better traction
at modelling cross-cultural variation through environmental statistics
such as lexical co-occurrences, rather than treating groups of cultures
as monoliths (East Asian vs Western). However, there are still many open
questions for this account. Our operationalization of environmental
statistics with lexical co-occurrence is a proxy of both language inputs
and culture-specific conceptualization of similarity. The extent to
which cross-cultural differences in similarity judgment are driven by
language versus culture is still an open question. Future work should
also aim to provide a more specific computational account of how lexical
co-occurrence might guide categorization preference beyond the simple
proportion-of-similarity model tested here, and investigate other
individual- or triad-specific factors that might influence how
environmental statistics affect similarity judgment.

@@
